---
title: "Analysis of the NOAA Storm Database and study the influence of the storm on health and economy"
output: html_document
author: Dmitry
---

##Synopsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.
In this document storms and other severe weather events from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database are analyzed. Using whole data from 1950 to November 2011 questions about severe weather events were studied. The most harmful type of events with respect to population health and type of events with the greatest economic consequences were determined.

There was found, that:
  
* Tornado is most harmful with respect to population health and cause about 
3.7% of all fatalities from weather events and 6.5% of all injuries from weather events
  
* Flood have the greatest economic consequences and cause 3.8% of all money losses from weather events

##Data Processing
For study was used this version of database file:
https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2 and description: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf

###Getting data
At the first, the datafile was downloaded and read.

```{r, cache=TRUE}
DataUrl <-  "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
fileName <- paste0(getwd(), "/../StormData.csv.bz2")
if(!file.exists(fileName)){
    download.file(DataUrl, fileName)
    }
NOAA_data <- read.csv(bzfile(fileName))
```

### Structure of data
There are several first lines of data presented to see structure of the data
```{r}
head(NOAA_data)
```

Our study should answer two queston:
  
* Across the United States, which types of events (as indicated in the `EVTYPE` variable) are most harmful with respect to population health?
  
* Across the United States, which types of events have the greatest economic consequences?

Therefore for this study will be useful next column: `EVTYPE` `FATALITIES` 
`INJURIES` `PROPDMG` `CROPDMG` `PROPDMGEXP` `CROPDMGEXP`


`PROPDMGEXP` `CROPDMGEXP` indicate, who damage could be calculated. There are severel types of data in these column.

```{r}
NOAA_data$PROPDMGEXP <- as.factor(NOAA_data$PROPDMGEXP)
NOAA_data$CROPDMGEXP <- as.factor(NOAA_data$CROPDMGEXP)

print(levels(NOAA_data$PROPDMGEXP))
print(levels(NOAA_data$CROPDMGEXP))

powIndicator <- levels(as.factor(c(levels(NOAA_data$PROPDMGEXP),
                                   levels(NOAA_data$CROPDMGEXP))))[-1:-4]
print(powIndicator)
```

Damage can be calculated as `Expense = A * 10^B`, where A is number from `CROPDMG`/`PROPDMG` field and B is exponential values from table. Numbers are simply B-value, `h/H` for hundreds,`k/K` for thousands, etc.

```{r, results='asis'}
library("xtable")
ConvTable<-data.frame(DMGEXP = powIndicator, B = c(0,1,2,3,4,5,6,7,8,9,2,2,3,3,6,6))
xt <- xtable(ConvTable)
print(xt, type="html",include.rownames=FALSE)

```

For our purposes `PROPDMG` `CROPDMG` field were recalculated acording this table to compute the total losses.

```{r}

for (e in ConvTable$DMGEXP){
    NOAA_data$PROPDMG[NOAA_data$PROPDMGEXP == e] <-
        NOAA_data$PROPDMG[NOAA_data$PROPDMGEXP == e]* 10 ^ ConvTable$B[ConvTable$DMGEXP == e]
    
    NOAA_data$CROPDMG[NOAA_data$CROPDMGEXP == e] <-
        NOAA_data$CROPDMG[NOAA_data$CROPDMGEXP == e]* 10 ^ ConvTable$B[ConvTable$DMGEXP == e]
}

```


##Results
 
###Analysis of weather events with respect to population health

There are aggregated harmful weather events with respect to population health.

```{r, cache=TRUE}
inj.sum <- aggregate(INJURIES ~ EVTYPE, NOAA_data, sum)
fat.sum <- aggregate(FATALITIES ~ EVTYPE, NOAA_data, sum)

```

There are 5 worst types of weather events with respect to population health.

* INJURIES

```{r}
inj.ten <- head(inj.sum[order(-inj.sum$INJURIES), ], 5)
print(inj.ten)
```

* FATALITIES

```{r}
fat.ten <- head(fat.sum[order(-fat.sum$FATALITIES), ], 5)
print(fat.ten)
```

So, the most harmful effect is **Tornado**. The most harmful effects in order to injuries and 
fatalities are plotted.

```{r}

par(mfrow=c(1,2), mar=c(8,3,4,2)+0.1)
barplot(inj.ten$INJURIES, main="Injuries", ylab="Numbers of injuries",
        las=3, names.arg = inj.ten$EVTYPE)

barplot(fat.ten$FATALITIES, main="Fatalities", ylab="Numbers of fatalities",
        las=2, names.arg = inj.ten$EVTYPE)
```

Finally, there is evaluation the percentage of injuries and fatalities from **Tornado**.

```{r}
h <- fat.ten$FATALITIES[1]/ sum(fat.sum$FATALITIES)
print(paste0("Percentage of fatalities from Tornado:", h * 10, " %"))

h <- inj.ten$INJURIES[1]/ sum(inj.sum$INJURIES)
print(paste0("Percentage of injuries from Tornado:", h * 10, " %"))
```
	



###Analysis of weather events with respect to economic consequences


Losses data for each type of damage was aggregated. 
```{r}
crop.sum <- aggregate(CROPDMG ~ EVTYPE, NOAA_data, sum)
prop.sum <- aggregate(PROPDMG ~ EVTYPE, NOAA_data, sum)
```

`PROPDMG` `CROPDMG` were summed, because they both mesured in US dollars. The names of most harmful events were extracted from each column (in variable `EventNames`)

```{r}

a <- head(crop.sum[order(-crop.sum$CROPDMG), ], 5)
b <- head(prop.sum[order(-prop.sum$PROPDMG), ], 5)
EventNames <- c(as.vector(a$EVTYPE), as.vector(b$EVTYPE))
crop.sum <- crop.sum[crop.sum$EVTYPE %in% EventNames, ]
prop.sum <- prop.sum[prop.sum$EVTYPE %in% EventNames, ]
p.sum <- merge(crop.sum, prop.sum, by="EVTYPE")
p.sum$SUM <- p.sum$CROPDMG + p.sum$PROPDMG

```

There are 5 worst types of weather events with respect to money losses.

```{r}
p.ten <- head(p.sum[order(-p.sum$SUM), ], 5)

row.names(p.ten) <- p.ten$EVTYPE
print(p.ten[,c(1,4)], row.names = FALSE)
```

So, the most harmful effect is **Flood**. The most harmful effects in order to money loss are plotted.

```{r}

par(mar=c(10,5,4,2)+0.3)
barplot(p.ten$SUM/ 1e9, main="Economic damage", ylab="Billion $",
        las=3, names.arg = p.ten$EVTYPE)
```

Finally, there is evaluation the percentage of money loss from **Flood**.

```{r}
h <- p.ten$SUM[1]/ (sum(crop.sum$CROPDMG) + sum(prop.sum$PROPDMG))
print(paste0("Percentage of money loss from Tornado:", h * 10, " %"))
```
	

