---
title: "Analysis of the NOAA Storm Database and study the influence of the storm on health and economy"
output: html_document
author: Dmitry
---

##Synopsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.
In this document storms and other severe weather events from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database are analyzed. Using whole data from 1950 to November 2011 questions about severe weather events were studied. The most harmful type of events with respect to population health and type of events with the greatest economic consequences were determined.
For study was used this version pf database file:
https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2 and description: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf


##Data Processing
For analysis this reference was used:
https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2

At the first, I download and read file

```{r}
DataUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
setwd("E:/Cloud/Coursera/DataScience/5_ReprResearch/PeerAssessment5.2")
download.file(DataUrl, destfile = "StormData.csv.bz2", method = "curl")
#readcsvbz2file <- read.csv(bzfile(DataUrl))

```

##Results
 



You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
